---
title: "esm244-w2021-lab3 part 1"
author: "Kelsie Fronheiser"
date: "2/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
library(GGally)
library(broom)
library(jtools)
```

```{r}
penguins %>% 
  select(species, bill_length_mm:sex) %>% 
  ggpairs(aes(color = species))
```

We are going to be looking at Adelies and Chinstraps because they are so similar in size
```{r}
adelie_chinstrap <- penguins %>% 
  filter(species %in% c ("Adelie", "Chinstrap")) %>% 
  mutate(species = fct_drop(species))
```

But Gentoo factor level is still within the data set so mutate(new name = fct_drop(existing species level)) - drops that factor
Check factor by typing "levels(adelie_chinstrap$species)" into console
Pay attention to order factors show up in

```{r}
#Data visualization

ggplot(data = adelie_chinstrap, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = sex)) +
facet_wrap(~species)
```

Can we capture the difference between the two?

Binomial logistic regression: classifying factors into "bins"

using variables: body mass, flipper length, penguin sex
```{r}
adelie_chinstrap_blr <- glm(species ~ body_mass_g + flipper_length_mm + sex,
                            data = adelie_chinstrap,
                            family = "binomial")
```

Take a look at outcomes
```{r}
adelie_chinstrap_blr

summary(adelie_chinstrap_blr)
```
What this means:
levels shows us that adelie is coded as 0 (first) and chinstrap is 1, so that meansin the log linear expression the left hand side is probability of chinstrap (log odds of non-zero level)

Body mass -8.576e_4: Expect on avg that as body mass increases the chances of being chinstrap may be going down

Flipper length 1.778e-1: Epect on avg that increase in flipper length the 
odds of it being a chinstrap penguin is increasing

```{r}
ggplot(data = adelie_chinstrap, aes(x = species, y = flipper_length_mm)) +
  geom_jitter(aes(color = sex))
```


Turning log probabilities into probabilities: broom function
What are the actual probabilities of being chinstrap penguin for each of the existing observations in our adelie_chinstrap data?
```{r}
blr_fitted <- adelie_chinstrap_blr %>% 
  broom::augment(type.predict = "response")
```

This adds column to dataset; the "fitted" column is the probability of being chinstrap according to each indivdual's variables (body mass, flipper length, sex)

You can see the probabilities are higher when the species is ineed chinstrap but still mistakes (some chinstraps low prob, some adelie high prob); how often would a penguin be correctly identified using this model?

Plotting fitted probabilities:
```{r}
ggplot(data = blr_fitted, aes(x = flipper_length_mm, y = .fitted)) +
  geom_point(aes(color = sex, pch = species)) +
  geom_smooth(aes(color = sex), se = FALSE) +
  labs(x = "Flipper length (mm)",
       y = "Probability of outcome Chinstrap")
```

As flipper length increases for both male and female penguins the prob of that penguin being Chinstrap is increasing

Effect plot function from jtools: provides us with quick overview of what our model looks like based on different variables (how prob changes depedent on variable)
```{r}
effect_plot(adelie_chinstrap_blr,
            pred = flipper_length_mm,
            y.label = "Probability of Chinstrap",
            interval = TRUE)


effect_plot(adelie_chinstrap_blr,
            pred = body_mass_g,
            y.label = "Probability of Chinstrap",
            interval = TRUE)
```



What's the probability that a female penguin weighing 3410g with a flipper length of 192mm is a Chinstrap?
```{r}
ex_1 <- predict(adelie_chinstrap_blr,
                data.frame(sex = "female",
                           body_mass_g = 3410,
                           flipper_length_mm = 192),
                type = "response")

ex_1

#but this is in log odds so add last line type = "response" to get in normal probability 
```


